---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "50%",
  fig.align = "center",
  dpi = 96
)
```

# gghotelling

<!-- badges: start -->
[![R-CMD-check](https://github.com/january3/gghotelling/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/january3/gghotelling/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/january3/gghotelling/graph/badge.svg)](https://app.codecov.io/gh/january3/gghotelling)
<!-- badges: end -->

Show Hotelling T² ellipses with ggplot2.

This is different from the `stat_ellipse()` function in ggplot2, which
draws normal data ellipses. Hotelling ellipses are multivariate confidence
regions for the mean of a multivariate normal distribution. This is also
different from the ellipses returned by the `ellipse::ellipse()` function, which
are confidence regions for the covariance matrix of a multivariate normal
distribution (data ellipses again).


## Installation

You can install the development version of gghotelling from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("january3/gghotelling")
```

## Example

The package defines a new geom, `geom_hotelling()`, which can be used to add Hotelling ellipses to ggplot2 scatter plots.

```{r example,fig.width=6, fig.height=4}
library(ggplot2)
library(gghotelling)

pca <- prcomp(iris[, 1:4], scale.=TRUE)
df <- cbind(iris, pca$x)

ggplot(df, aes(PC1, PC2)) +
  geom_hotelling(ci=.99) +
  geom_point()

ggplot(df, aes(PC1, PC2, color=Species)) +
  geom_hotelling() +
  geom_point()

ggplot(df, aes(PC1, PC2, color=Species)) +
  geom_hotelling(alpha=0.1, aes(fill = Species)) +
  geom_point()
```


