---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "50%",
  fig.align = "center",
  fig.width = 8,
  fig.height = 5,
  dpi = 96
)
```

# Hotelling's T² Ellipses and Outlier Detection for ggplot2

<!-- badges: start -->
[![R-CMD-check](https://github.com/january3/gghotelling/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/january3/gghotelling/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/january3/gghotelling/graph/badge.svg)](https://app.codecov.io/gh/january3/gghotelling)
<!-- badges: end -->

Hotelling data ellipses use the Hotelling T² distribution to create
coverage regions for the distribution of the data, often used in outlier
detection in multivariate data. This is different from the ellipses
returned by the `ellipse::ellipse()` or `car::dataEllipse()` functions,
which produce data ellipses based on a Mahalanobis distance contour,
representing the shape and spread of the data distribution.

This is also slightly different from the `stat_ellipse()` can be used to
create data ellipses (much like the ones from `ellipse::ellipse()`), and
also certain type of t ellipses. In contrast, gghotelling provides
explicit Hotelling T² data ellipses and Hotelling
T² confidence ellipses, with a clear distinction between the two.

Unlike stat_ellipse(), it can also take
the `fill` aesthetic for a visually pleasing representation of the ellipses.

A lot of functionality overlaps with `ggfortify::ggbiplot()` (and by
extension `autoplot.pca_class`),
but this function is less flexible than a separate geom that you can add
to the figure.

Moreover, the package provides a `stat_hotelling_points()` function to compute
the per-point Hotelling T² statistics, which can be used to identify and
visualize multivariate outliers in scatter plots.

My main motivation for creating this package was sorting out the different
ellipse types and allowing the use of `fill` aesthetics for Hotelling ellipses.

## Installation

You can install the development version of gghotelling from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("january3/gghotelling")
```

## Hotelling Ellipses

The package defines a new geom, `geom_hotelling()`, which can be used to add Hotelling ellipses to ggplot2 scatter plots.

```{r example}
library(ggplot2)
library(gghotelling)

pca <- prcomp(iris[, 1:4], scale.=TRUE)
df <- cbind(iris, pca$x)

ggplot(df, aes(PC1, PC2)) +
  geom_hotelling(level=.99) +
  geom_point()

ggplot(df, aes(PC1, PC2, color=Species)) +
  geom_hotelling() +
  geom_point()

ggplot(df, aes(PC1, PC2, color=Species)) +
  geom_hotelling(alpha=0.1, aes(fill = Species)) +
  geom_point()

# set custom CI/coverage level
ggplot(df, aes(PC1, PC2, color=Species)) +
  geom_hotelling(alpha=0.1, aes(fill = Species), level=.99) +
  geom_point()
```

The package also provides per-point, group-wise T² statistics which can be
used to identify multivariate outliers.



```{r example2}
ggplot(df, aes(PC1, PC2, group=Species)) +
  geom_hotelling(level = 0.75, alpha=0.1, aes(fill = Species)) +
  scale_color_manual(values=c("TRUE"="red", "FALSE"="grey")) +
  stat_hotelling_points(level = .75, aes(shape = Species, color = after_stat(is_outlier)))

ggplot(df, aes(PC1, PC2, group=Species)) +
  geom_hotelling(alpha=0.1, aes(fill = Species)) +
  stat_hotelling_points(size=2, aes(shape = Species, color = after_stat(t2)))
```

This can be useful for identifying potential outliers in multivariate data.
The outliers can be directly labeled as follows:


```{r example3}
ggplot(df, aes(PC1, PC2, group=Species, label=rownames(df))) +
  geom_hotelling(alpha=0.1, aes(fill = Species)) +
  geom_point(aes(color = Species)) +
  stat_hotelling_points(geom="label", 
                        outlier_only = TRUE)

# or even with geom_label_repel from ggrepel
library(ggrepel)
ggplot(df, aes(PC1, PC2, group=Species, label=rownames(df))) +
  geom_hotelling(alpha=0.1, aes(fill = Species)) +
  geom_point(aes(color = Species)) +
  stat_hotelling_points(geom="label_repel",
                        outlier_only = TRUE)
```

## Convex Hulls


```{r example_hull}
ggplot(iris, aes(Sepal.Length, Sepal.Width, color=Species)) +
  geom_hull(mapping = aes(fill = Species), alpha=.1) +
  geom_point()
```




## Autoplot

The package also defines `autoplot.prcomp` and `autolayer.prcomp` for
convenient plotting of PCA plots. Note that `autoplot.prcomp` is also
implemented in a more sophisticated way in the `ggfortify` package.


```{r example_autoplot}
autoplot(pca, group = iris$Species) + 
  autolayer(pca, group = iris$Species)
```




