---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "50%",
  fig.align = "center",
  dpi = 96
)
```

# gghotelling

<!-- badges: start -->
[![R-CMD-check](https://github.com/january3/gghotelling/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/january3/gghotelling/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/january3/gghotelling/graph/badge.svg)](https://app.codecov.io/gh/january3/gghotelling)
<!-- badges: end -->

Show Hotelling T² ellipses with ggplot2.

Hotelling data ellipses are using the Hotelling T² distribution to create
coverage regions for the distribution of the data, often used in outlier
detection in multivariate data. This is different from the ellipses
returned by the `ellipse::ellipse()` or `car::dataEllipse()` functions,
which produce data ellipses based on a χ² (chi-square) contour,
representing the shape and spread of the data distribution.

This is also slightly different from the `stat_ellipse()` can be used to
create data ellipses (much like the ones from `ellipse::ellipse()`), and
also Hotelling data (?) ellipses. Unlike stat_ellipse(), it can also take
the `fill` aesthetic for a visually pleasing representation of the ellipses.

My main motivation for creating this package was sorting out the different
ellipse types and allowing the use of `fill` aesthetics for Hotelling ellipses.

## Installation

You can install the development version of gghotelling from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("january3/gghotelling")
```

## Example

The package defines a new geom, `geom_hotelling()`, which can be used to add Hotelling ellipses to ggplot2 scatter plots.

```{r example,fig.width=10, fig.height=5}
library(ggplot2)
library(gghotelling)

pca <- prcomp(iris[, 1:4], scale.=TRUE)
df <- cbind(iris, pca$x)

ggplot(df, aes(PC1, PC2)) +
  geom_hotelling(ci=.99) +
  geom_point()

ggplot(df, aes(PC1, PC2, color=Species)) +
  geom_hotelling() +
  geom_point()

ggplot(df, aes(PC1, PC2, color=Species)) +
  geom_hotelling(alpha=0.1, aes(fill = Species)) +
  geom_point()

# set custom CI/coverage level
ggplot(df, aes(PC1, PC2, color=Species)) +
  geom_hotelling(alpha=0.1, aes(fill = Species), ci=.99) +
  geom_point()
```

The package also provides per-point, group-wise T² statistics which can be
used to identify multivariate outliers.



```{r example2,fig.width=10, fig.height=5}
ggplot(df, aes(PC1, PC2, group=Species)) +
  geom_hotelling(ci = 0.75, alpha=0.1, aes(fill = Species)) +
  scale_color_manual(values=c("TRUE"="red", "FALSE"="grey")) +
  stat_hotelling_points(ci = .75, aes(shape = Species, color = after_stat(outside)))

ggplot(df, aes(PC1, PC2, group=Species)) +
  geom_hotelling(alpha=0.1, aes(fill = Species)) +
  stat_hotelling_points(size=2, aes(shape = Species, color = after_stat(t2)))
```

This can be useful for identifying potential outliers in multivariate data.
The outliers can be directly labeled as follows:


```{r}
ggplot(df, aes(PC1, PC2, group=Species, label=rownames(df))) +
  geom_hotelling(alpha=0.1, aes(fill = Species)) +
  geom_point(aes(color = Species)) +
  stat_hotelling_points(geom="label", 
                        outlier_only = TRUE)

# or even with geom_label_repel from ggrepel
library(ggrepel)
ggplot(df, aes(PC1, PC2, group=Species, label=rownames(df))) +
  geom_hotelling(alpha=0.1, aes(fill = Species)) +
  geom_point(aes(color = Species)) +
  stat_hotelling_points(geom="label_repel",
                        outlier_only = TRUE)
```




